---
title: "Read and Merge bilingual CDI Data"
author: "Krista Byers-Heinlein & Hilary Killam based on script from Esther Schott"
output: html_notebook
---

#Prep
## Load packages
```{r}
library(knitr)
library(tidyverse)
library(here)
library(readxl)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(janitor)
library(gvlma)
library(performance)
library(sjPlot)
library(tidylog)
options(scipen = 999)

```

## Read in data
```{r}

# Words and Sentences dataset- older children
cdi_ws_dataset <- read_csv(here("data/byers-heinlein_ws_summary_data_en-fr.csv"))

cdi_ws <- cdi_ws_dataset %>% # Create new dataset to work with
  rename(word_production = word_vocab_production) %>%
  rename(concept_production = concept_vocab_production) %>%
  rename(lang_production = lang_vocab_production) %>%
  mutate(word_comprehension = NA) %>%
  mutate(concept_comprehension = NA) %>%
  mutate(te_comprehension = NA) 
  
# Words and Gestures dataset - younger children

cdi_wg_dataset <- read_csv(here("data/byers-heinlein_wg_summary_data_en-fr.csv"))

cdi_wg <- cdi_wg_dataset %>% # Create new dataset to work with
  rename(word_production = word_vocab_production) %>%
  rename(word_comprehension = word_vocab_comprehension) %>%
  rename(concept_production = concept_vocab_production) %>%
  rename(concept_comprehension = concept_vocab_comprehension) %>% 
  rename(lang_production = lang_vocab_production) %>%
  rename(lang_comprehension = lang_vocab_comprehension)

```
## Combine wg and ws datasets
```{r}

cdi <- full_join(cdi_wg, cdi_ws) %>%
  mutate(form_id = paste(form_id, form))

```

## Language group - select monolinguals and French/English bilinguals
```{r}

# Determine language group
cdi <- cdi %>%
  mutate(lang_group = case_when(
  exposure_to_english >= .90 ~ "monolingual", # Monolingual >90% English or French
  exposure_to_french >= .90 ~ "monolingual",
  exposure_to_english >= .25 & exposure_to_french >= .25 ~ "bilingual", # Bilingual at least 25% English and French
  exposure_to_other_language >= .10 ~ "other", # No third language >10%
  TRUE ~ "other")) 

n_other <- cdi %>% #421 other-language children
  filter(lang_group == "other") %>%
  summarize(n_other = n_distinct(child_id)) %>%
  pull(n_other)

#How many administrations in each lang group?
cdi %>% 
  count(lang_group)

# Filter out "other"

cdi <- cdi %>%
  filter(lang_group != "other") %>%
  mutate(lang_group = as.factor(lang_group))

```


## Remove premature, low birthweight, age
```{r}

premature <- cdi %>% count(premature) # 48 with no information
health <- cdi %>% count(health_issues) # 57 no information
birthweight <- cdi %>% count(low_birthweight) # 124 no information
wrong_age <- cdi %>% count(child_age_in_months) # 2 7-month-olds, will include the 12 17-month-olds


cdi <- cdi %>%
  filter(premature == "no") %>%
  filter(health_issues == "no") %>%
  filter(low_birthweight == "no") %>%
  filter(child_age_in_months >= 8)

n_premature_health_age_removed <- cdi %>% # 948 distinct visits
  summarize(n_visits = n_distinct(visit_id)) %>%
  pull(n_visits)


```
## Dominance 
```{r dominance}

# Determine which language is dominant

cdi <- cdi %>%
  mutate(dom_lang = case_when(
    exposure_to_english > exposure_to_french ~ "english",
    exposure_to_french > exposure_to_english ~ "french",
    exposure_to_english == exposure_to_french ~ "balanced", 
    exposure_to_english >= .90 ~ "english", # to handle a few NA
    exposure_to_french >= .90 ~ "french"
  ))

# For balanced kids, make lang with higher vocab score dominant. See if consistent across production/comprehension:

cdi %>%
  filter(dom_lang == "balanced") %>%
  select(child_id, visit_id, exposure_to_english, exposure_to_french, lang_production, lang_comprehension, lang_group, form_language) %>%
  mutate(form_language = str_to_lower(str_remove(form_language, " \\(.*\\)"))) %>%
  pivot_wider(names_from = "form_language", values_from = c("lang_production", "lang_comprehension")) %>%
  mutate(domlangprod = case_when(lang_production_english > lang_production_french ~ "en",
                                 lang_production_french > lang_production_english ~ "fr",
                                 is.na(lang_production_english) & !is.na(lang_production_french) ~ "fr",
                                 TRUE ~ "check"),
         domlangcomp = case_when(lang_comprehension_english > lang_comprehension_french ~ "en",
                                 lang_comprehension_french > lang_comprehension_english ~ "fr",
                                 is.na(lang_comprehension_english) & is.na(lang_comprehension_french) ~ domlangprod,
                                 TRUE ~ "check"),
         domlangprod = case_when(lang_production_english == lang_production_french ~ domlangcomp,
                                 TRUE ~ domlangprod),
         balance_check = ifelse(domlangcomp == domlangprod, "same", "DIFFERENT")) %>%  
  arrange(child_id)

#One kid with equal exposure to FR and EN has higher productive vocab in English but higher receptive vocab in French. We will go with higher productive vocab to assign dominance.

#How many kids have different lang dominances at different visits?
cdi %>%
  select(child_id, lab_visit_num, visit_id, exposure_to_english, exposure_to_french, lang_group, form_language, dom_lang) %>%
  group_by(child_id) %>%
  filter(n_distinct(dom_lang) > 1)%>%
  distinct(child_id) #looks like 18 kids switch language dominance between visits, mostly fairly balanced bilinguals.

#---------------------------------Set dom lang to higher productive score for those balanced kids, or if prod score is equal then comp score

fixdomlang <- cdi %>%
  filter(dom_lang == "balanced") %>%
  select(child_id, visit_id, exposure_to_english, exposure_to_french, lang_production, lang_comprehension, lang_group, form_language) %>%
  mutate(form_language = str_to_lower(str_remove(form_language, " \\(.*\\)"))) %>%
  pivot_wider(names_from = "form_language", values_from = c("lang_production", "lang_comprehension")) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0))) %>% #make NAs 0 temporarily to calculate dominance
  mutate(dom_lang = case_when(
    exposure_to_english == exposure_to_french & lang_production_english > lang_production_french ~ "english",
    exposure_to_english == exposure_to_french & lang_production_french > lang_production_english ~ "french",
    exposure_to_english == exposure_to_french & lang_comprehension_english > lang_comprehension_french ~ "english",
    exposure_to_english == exposure_to_french & lang_comprehension_french > lang_comprehension_english ~ "french"
  )) %>%
  select(child_id, visit_id, dom_lang_fixed = dom_lang)

cdi <- cdi %>%
  left_join(fixdomlang) %>%
  mutate(dom_lang = case_when(dom_lang == "balanced" ~ dom_lang_fixed,
                              TRUE ~ dom_lang)) %>%
  select(-dom_lang_fixed)

cdi %>% count(dom_lang) # 785 English dominant, 587 French_dominant administrations

cdi <- cdi %>%
  mutate(dom_cdi = case_when(
    dom_lang == "english" & form_language == "English (American)" ~ "dominant",
    dom_lang == "english" & form_language == "French (Quebecois)" ~ "nondominant",
    dom_lang == "french" & form_language == "French (Quebecois)" ~ "dominant",
    dom_lang == "french" & form_language == "English (American)" ~ "nondominant",
    dom_lang == "balanced" ~ "balanced"
  ))

cdi %>% count(dom_cdi) # 936 administrations in dominant language, 434 in non-dominant language

```

## For monolinguals remove non-dominant language and remove other lang scores from concept and productive vocab. Set mono TEs to 0
```{r}

cdi <- cdi %>%
  filter(!(lang_group == "monolingual" & dom_cdi == "nondominant")) %>% # for monolinguals, filter if not dominant lang vocab
  #Hbecause we are removing 2nd CDIs for monolinguals, should recode values for "forms_filled_visit" to be consistent with the data
  mutate(forms_filled_visit = case_when(lang_group == "monolingual" & forms_filled_visit == "English & French" ~ paste0(str_to_title(dom_lang), " only"),
                                        TRUE ~ forms_filled_visit)) %>%
  mutate(concept_production = case_when(
    lang_group == "monolingual" ~ lang_production,
    TRUE ~ concept_production
  )) %>%
  mutate(word_production = case_when(
    lang_group == "monolingual" ~ lang_production,
    TRUE ~ word_production
  )) %>%
  mutate(te_production = case_when(
    lang_group == "monolingual" ~ 0,
    TRUE ~ te_production
  ))
```

# Participant characteristics

```{r}


# n administrations, 1338 administrations
cdi %>%
  summarize(n = n_distinct(form_id))

#n administrations per lang group, 510 monolingual, 828 bilingual
cdi %>%
  group_by(lang_group) %>%
  summarize(n = n_distinct(form_id))

#n administrations per lang group and dom lang, 472 En-dom bilingual administrations, 356 Fr-dom bi, 277 En mono, 233 Fr mono
cdi %>%
  group_by(lang_group, dom_lang) %>%
  summarize(n = n_distinct(form_id)) %>%
  group_by(lang_group) %>%
  mutate(pct = n/sum(n))
  
#n children = 743
cdi %>%
  summarize(n = n_distinct(child_id))

#n children per lang group, 337 monolingual, 416 bilingual. Note: 10 kids switch lang group between visits
cdi %>%
  group_by(lang_group) %>%
  summarize(n = n_distinct(child_id))

#10 kids switch lang group between visits
cdi %>%
  group_by(child_id) %>%
  filter(n_distinct(lang_group) > 1) %>%
  distinct(child_id) %>%
  nrow()

#n visits per lang group, 437 bilingual visits, 510 monolingual visits
cdi %>%
  group_by(lang_group) %>%
  summarize(n = n_distinct(visit_id))

#n children per dom lang per lang group, 196 En-dom bilinguals, 157 Fr-dom bilinguals, 217 En monos, 199 Fr monos
cdi %>%
  group_by(lang_group, dom_lang) %>%
  summarize(n = n_distinct(child_id)) 

#n visits per dominant language, 523 En-dom visits, 424 Fr-dom visits
cdi %>%
  group_by(dom_lang) %>%
  summarize(n = n_distinct(visit_id)) 

# total ns for comprehension, 163 bi 253  mono
cdi %>%
  group_by(lang_group) %>%
  filter(!is.na(word_comprehension)) %>% #filter out WS forms which don't have comprehension scores
  summarize(n = n_distinct(child_id)) 

# total ns for production, 337 bi 416 mono
cdi %>%
  group_by(lang_group) %>%
  summarize(n = n_distinct(child_id)) 

# parental education

 #percent of visits with education info and which parent: 98.5% mothers, 1.2% not reported, 0.3% fathers
cdi %>% 
  distinct(visit_id, edu_parent) %>% 
  tabyl(edu_parent)

 #average years of education by lang group, bi = 16.9 mono = 16.7
cdi %>% 
  distinct(visit_id, lang_group, years_education) %>% 
  group_by(lang_group) %>%
  summarize(av_yrs_edu = mean(years_education, na.rm = TRUE)) 

 #highest edu level, 3 mothers of monolinguals have high school, everyone else has at least some postsecondary
cdi %>%
  distinct(visit_id, highest_edu_desc, edu_parent, lang_group) %>%
  count(highest_edu_desc, edu_parent, lang_group) 

#gender, 350 girsl 393 boys (47% female)
cdi %>% 
  distinct(child_id, sex) %>%
  tabyl(sex)
 
#n kids per gender per lang group, 156 bilingual girls, 181 bilingual boys, 195 mono girls, 221 mono boys
cdi %>% 
  distinct(child_id, lang_group, sex) %>%
  count(lang_group, sex) %>%
  group_by(lang_group) %>%
  mutate(pct = n/sum(n))

#who filled out CDI
cdi %>% 
  distinct(form_id,responding_caregiver, lang_group) %>%
  count(lang_group, responding_caregiver) %>%
  group_by(lang_group) %>%
  mutate(pct = n/sum(n)) %>%
  arrange(lang_group, desc(n))

#which CDI versions? 647 WG administrations, 691 WS administrations
cdi %>%
  count(form)

```



# Clean data ready for regression

##Centre age
```{r}
# Create variables to center age at 12 months for comprehension models
# make monolinguals the reference df
cdi <- cdi %>%
  mutate(lang_group = fct_relevel(lang_group, "monolingual", "bilingual")) %>%
  mutate(age_decimal_months = child_age_in_days/30.41667) %>% #use age in days and divide by average days in a month to get their age in months with a decimal, for easier plot visualization. 30.41667 is 365/12
  mutate(age_decimal_months_centred = age_decimal_months - 12)

```


## Make dataset long

```{r}

# Pivot word and concept vocabs, only keep values if both forms are filled out
cdi_total_scores <- cdi %>%
  select(-form_language, -responding_caregiver) %>%
  pivot_longer(cols = c("word_production", "concept_production", "lang_production", "word_comprehension", "concept_comprehension", "lang_comprehension", "te_comprehension", "te_production"), names_to = c("vocab_measure", "vocab_modality"), values_to = "vocab_score", names_sep = "_") %>%
  filter(lang_group == "monolingual" | forms_filled_visit == "English & French") %>%
  select(-form_id) %>%
  filter(!is.na(vocab_score)) %>%    # remove NAs
  mutate(vocab_measure = case_when(
    vocab_measure == "lang" ~ dom_cdi, # indicate whether the score is for dom or nondom language
    TRUE ~ vocab_measure)) %>%
  select(-dom_cdi) %>% # don't need dom_cdi column anymore
  distinct()  # removes duplicate rows


# Create non-dominant for monolinguals (this is the same as the value for dominant)
cdi_mono_nondom <- cdi_total_scores %>%
  filter(lang_group == "monolingual" & vocab_measure == "dominant") %>%
  mutate(vocab_measure = "nondominant")
  

# Join total scores with monolingual dominant
cdi_long <- cdi_total_scores %>%
  full_join(cdi_mono_nondom) %>%
  relocate(vocab_score, .after = last_col()) %>%
  mutate(vocab_measure = fct_relevel(vocab_measure, "word", "concept", "dominant", "nondominant")) #reorder factor levels


```

## Check number of cases for dominant, non-dominant, word, and concept
```{r}

cdi_long %>%
  group_by(lang_group, vocab_measure, vocab_modality) %>%
  tally()

```

# DATA VISUALIZATION

## Visualize production - points above 750 not plotted
```{r}
#Note: this model is quadratic, use polynomial formula for geom_smooth
cdi_long %>%
  filter(vocab_modality == "production") %>%
  filter(vocab_measure != "balanced") %>%
  filter(vocab_measure != "te") %>%
  ggplot(aes(x = age_decimal_months, y = vocab_score, color = lang_group, fill = lang_group, lty = lang_group)) +
  geom_point(alpha = 0.5, shape = 21, stroke = NA, color = "#FFFFFF") +
  #  stat_smooth(se = FALSE, span = 4) + # adjusted smoothing parameter
  stat_smooth(se = FALSE, formula = y ~ x + I(x^2), method = "lm") +
  facet_grid(.~vocab_measure, scales = "free") +
  coord_cartesian(ylim = c(0,750)) +
  ggtitle("Production") +
  theme_bw(base_size = 12) +
  scale_color_manual(name="language group",
                          values =c("#377eb8", "#e41a1c")) +
  scale_fill_manual(name="language group",
                          values =c("#c9e7ff", "#ffc6c6")) +
  scale_linetype_manual(name = "language group", values = c("solid", "dashed")) +
  theme (legend.position = "bottom", legend.key.size = unit(3,"line")) +
   xlab("child age in months") +
  ylab("vocabulary size")
  
  ggsave(here("./figures/production.png"), dpi=300, height=4, width=7, units="in")

```


## Visualize comprehension

```{r}
#Note: this model is linear, use linear formula for geom_smooth
cdi_long %>%
  filter(vocab_modality == "comprehension") %>%
  filter(vocab_measure != "balanced") %>%
  filter(vocab_measure != "te") %>%
  ggplot(aes(x = age_decimal_months, y = vocab_score, color = lang_group, fill = lang_group, lty = lang_group)) +
  geom_point(alpha = 0.5, shape = 21, stroke = NA, color = "#FFFFFF") +
  #  stat_smooth(se = FALSE, span = 4) + # adjusted smoothing parameter
  stat_smooth(se = FALSE, formula = y ~ x, method = "lm") +
  facet_grid(.~vocab_measure, scales = "free") +
  coord_cartesian(ylim = c(0,400)) +
  ggtitle("Comprehension") +
  theme_bw(base_size = 12) +
  scale_color_manual(name="language group",
                          values =c("#377eb8", "#e41a1c")) +
  scale_fill_manual(name="language group",
                          values =c("#c9e7ff", "#ffc6c6")) +
  scale_linetype_manual(name = "language group", values = c("solid", "dashed")) +
  theme (legend.position = "bottom", legend.key.size = unit(3,"line")) +
   xlab("child age in months") +
  ylab("vocabulary size")

ggsave(here("./figures/comprehension.png"), dpi=300, height=4, width=7, units="in")


```

#MODELS
## Production models
```{r}

#-------------------------------------------------BASE MODELS - LINEAR FIT

prod_models_base <- cdi_long %>%
  filter(vocab_modality == "production") %>%
  group_by(vocab_measure) %>%
  nest() %>%
  mutate(base_model = map(data, function(df) lmer(vocab_score ~ age_decimal_months*lang_group + (1|child_id), data = df, REML = FALSE)))  %>%
  filter(vocab_measure != "te") #don't need this one

#-------------------------------------------------GROWTH CURVE MODELS - QUADRATIC FIT

prod_models_quadratic <- prod_models_base %>%
  mutate(quadratic_model = map(data, function(df) lmer(vocab_score ~ poly(age_decimal_months, 2)*lang_group + (1|child_id), data = df, REML = FALSE))) 

#-------------------------------------------------COMPARE MODELS

prod_models_comparison <- prod_models_quadratic %>%
  mutate(compare_models = map2(base_model, quadratic_model, stats::anova))

prod_models_comparison$compare_models #for production, all the quadratic models are a better fit of the data than linear models

```


## Comprehension models
```{r}

#-------------------------------------------------BASE MODELS - LINEAR FIT

comp_models_base <- cdi_long %>%
  filter(vocab_modality == "comprehension") %>%
  group_by(vocab_measure) %>%
  nest() %>%
  mutate(base_model = map(data, function(df) lmer(vocab_score ~ age_decimal_months*lang_group + (1|child_id), data = df, REML = FALSE))) %>%
  filter(vocab_measure != "te") #don't need this one

#-------------------------------------------------GROWTH CURVE MODELS - QUADRATIC FIT

comp_models_quadratic <- comp_models_base %>%
  mutate(quadratic_model = map(data, function(df) lmer(vocab_score ~ poly(age_decimal_months, 2)*lang_group + (1|child_id), data = df, REML = FALSE))) 

#-------------------------------------------------COMPARE MODELS

comp_models_comparison <- comp_models_quadratic %>%
  mutate(compare_models = map2(base_model, quadratic_model, stats::anova))

comp_models_comparison$compare_models #for comprehension, all the linear models are a better fit of the data than quadratic models

#-------------------------------------------------CENTRE BASE MODEL AGE TERM FOR FINAL MODEL

comp_models_linear_centred <- comp_models_base %>%
  mutate(linear_model = map(data, function(df) lmer(vocab_score ~ age_decimal_months_centred*lang_group + (1|child_id), data = df, REML = FALSE))) 

```
# Tables

```{r}

#-------------------------------------------Make Pretty Tables

#--------------Production

#check order of terms
prod_models_quadratic[1,] #1 = word
prod_models_quadratic[2,] #2 = concept
prod_models_quadratic[3,] #3 = dominant
prod_models_quadratic[4,] #4 = nondominan

tab_wc_prod <- sjPlot::tab_model(prod_models_quadratic$quadratic_model[[1]], #word
                                 prod_models_quadratic$quadratic_model[[2]], #concept
                    pred.labels = c("Intercept", "Age (Linear Term)", "Age (Quadratic Term)", 
                                    "Bilingual", "Age (Linear Term) * Bilingual", 
                                    "Age (Quadratic Term) * Bilingual"),
                    dv.labels = c("Word Vocabulary", "Concept Vocabulary"),
                    show.r2 = TRUE,
                    show.obs = FALSE,
                    show.re.var = FALSE,
                    show.icc = FALSE,
                    show.ngroups = FALSE,
                    title = "Expressive Vocabulary - Combined Measures")

tab_dn_prod <- sjPlot::tab_model(prod_models_quadratic$quadratic_model[[3]], #dominant
                                 prod_models_quadratic$quadratic_model[[4]], #nondominant
                    pred.labels = c("Intercept", "Age (Linear Term)", "Age (Quadratic Term)", 
                                    "Bilingual", "Age (Linear Term) * Bilingual", 
                                    "Age (Quadratic Term) * Bilingual"),
                    dv.labels = c("Dominant Vocabulary", "Non-Dominant Vocabulary"),
                    show.r2 = TRUE,
                    show.obs = FALSE,
                    show.re.var = FALSE,
                    show.icc = FALSE,
                    show.ngroups = FALSE,
                    title = "Expressive Vocabulary - Single-Language Measures")

#--------------Comprehension

#check order of terms
comp_models_linear_centred[1,] #word
comp_models_linear_centred[2,] #concept
comp_models_linear_centred[3,] #dominant
comp_models_linear_centred[4,] #nondominant

tab_wc_comp <- sjPlot::tab_model(comp_models_linear_centred$linear_model[[1]], comp_models_linear_centred$linear_model[[2]],
                    pred.labels = c("Intercept", "Age", "Bilingual", 
                                    "Age * Bilingual"),
                    dv.labels = c("Word Vocabulary", "Concept Vocabulary"),
                    show.r2 = TRUE,
                    show.obs = FALSE,
                    show.re.var = FALSE,
                    show.icc = FALSE,
                    show.ngroups = FALSE,
                    title = "Receptive Vocabulary - Combined Measures")

tab_dn_comp <- sjPlot::tab_model(comp_models_linear_centred$linear_model[[3]], comp_models_linear_centred$linear_model[[4]],
                    pred.labels = c("Intercept", "Age", "Bilingual", 
                                    "Age * Bilingual"),
                    dv.labels = c("Dominant Vocabulary", "Non-Dominant Vocabulary"),
                    show.r2 = TRUE,
                    show.obs = FALSE,
                    show.re.var = FALSE,
                    show.icc = FALSE,
                    show.ngroups = FALSE,
                    title = "Receptive Vocabulary - Single-Language Measures")


```


# Check model assumptions


## Production Models
```{r prod-check}
#Word
check_model(prod_models_quadratic$quadratic_model[[1]]) #looks mostly okay, some curvature where should be flat but not egregious
#Concept
check_model(prod_models_quadratic$quadratic_model[[2]]) #same as above
#Dominant
check_model(prod_models_quadratic$quadratic_model[[3]]) #same as above
#Nondominant
check_model(prod_models_quadratic$quadratic_model[[4]]) #same

```
## Comprehension Models 
```{r comprehension-check}

#Word
check_model(comp_models_linear_centred$linear_model[[1]]) #modtly okay, some curvature at edges and a couple influential points
#Concept
check_model(comp_models_linear_centred$linear_model[[2]]) #Some influential observations here
#Dominant
check_model(comp_models_linear_centred$linear_model[[3]]) #same as above
#Nondominant
check_model(comp_models_linear_centred$linear_model[[4]]) #same as above

```
